Analyzing the Citation Rate of a Scientific Paper Using ML and Graph Algorithms

### 2024-12-17 paper/paper.tex
Improve existing paper Add figure captions and references in text, add comparison of clusterization algorithms and some formulas. Also add table of contents.

\section{Introduction}

@@

\subsection{Degree Distribution}
\tab One of the important aspects of any graph is the distribution of the degrees of nodes. Since our graph is directed, we will have 2 distributions: one for the in-degree (number of incoming edges) and one for the out-degree (number of outgoing edges).

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{degree_distribution.png}
\end{figure}

@@

\subsection{Centralities}
\tab Centralities are characteristics of nodes that determine how "important" or "popular" each of them is. There are several types of them and each measures this "importance" in its own way. We will visualize distribution of three centralities.
\begin{itemize}
\item[a)] Eigenvector centrality. In general, relative scores are assigned to all nodes in the network based on the concept that connections to high-scoring nodes contribute more to the score of the node in question than equal connections to low-scoring nodes.
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{centralities_eigenvector.png}
\end{figure}
\item[b)] Closeness centrality. It is calculated as the reciprocal of the sum of the length of the shortest paths between the node and all other nodes in the graph. It measures how "close" the node is to the rest of the graph.
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{centralities_closeness.png}
\end{figure}
\item[c)] Betweenness centrality. The betweenness centrality for each node is the number of shortest paths that pass through this node. In other words, it determines whether the node is a hub.
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{centralities_betweenness.png}
\end{figure}
\end{itemize}

\subsection{Time Properties}
\tab Since our graph can be interpreted as a temporal graph, we can study the way its properties change over time. For instance, we discovered that the number of new papers per unit of time decreases:
\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]{new_papers_over_time.png}
\end{figure}
\newpage
Another interesting thing that we found is how average in-degrees and out-degrees change. This is the plot of average degrees in 2 months time from 1991 to 2003:
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{degrees_over_time.png}
\end{figure}
A downwards trend in the end for in-degree and upwards trend in the beginning are noticeable. Explanation to this phenomena is rather simple: older papers had more time to gain popularity and therefore have more incoming edges, whereas papers from 2003 do not have any papers in our dataset that cite them. The opposite happens with the out-degree. Earlier papers cited even earlier papers (from 1980s) that were not included in our graph and thus have little to no outgoing edges. It is also worth noticing that the growth of out-degree slows down closer to 2003. It means that almost enough papers are included in the graph at that time to publish a paper sourcing only these modern papers in our dataset.

>>>>

\tableofcontents

\newpage
\section{Introduction}

@@

\subsection{Degree Distribution}
\tab One of the important aspects of any graph is the distribution of the degrees of nodes. Since our graph is directed, we will have 2 distributions: one for the in-degree (number of incoming edges) and one for the out-degree (number of outgoing edges). Both of their distributions in log-scale are show on Figure \ref{plot:degree}.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{degree_distribution.png}
\caption{Degrees distribution.}
\label{plot:degree}
\end{figure}

@@

\subsection{Centralities}
\tab Centralities are characteristics of nodes that determine how "important" or "popular" each of them is. There are several types of them and each measures this "importance" in its own way. We will visualize distribution of three centralities.
\begin{itemize}
\item[a)] Eigenvector centrality. In general, relative scores are assigned to all nodes in the network based on the concept that connections to high-scoring nodes contribute more to the score of the node in question than equal connections to low-scoring nodes.
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{centralities_eigenvector.png}
\caption{Eigenvector centrality distribution.} \label{plot:centrality:eigenvector}
\end{figure}
\item[b)] Closeness centrality. It is calculated as the reciprocal of the sum of the length of the shortest paths between the node and all other nodes in the graph. It measures how "close" the node is to the rest of the graph.
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{centralities_closeness.png}
\caption{Closeness centrality distribution.} \label{plot:centrality:closeness}
\end{figure}
\item[c)] Betweenness centrality. The betweenness centrality for each node is the number of shortest paths that pass through this node. In other words, it determines whether the node is a hub.
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{centralities_betweenness.png}
\caption{Eigenvector centrality distribution.} \label{plot:centrality:betweenness}
\end{figure}
\end{itemize}

\subsection{Time Properties}
\tab Since our graph can be interpreted as a temporal graph, we can study the way its properties change over time. For instance, we discovered that the number of new papers per unit of time increases but the rate of the increase decreases. See Figure \ref{plot:number_of_new_papers} for the plots.
\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]{new_papers_over_time.png}
\caption{Number of new papers published each two months and each year.}
\label{plot:number_of_new_papers}
\end{figure}

Another interesting thing that we found is how average in-degrees and out-degrees change (see Figure \ref{plot:average_degree_per_month}).
\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{degrees_over_time.png}
\caption{Average degrees in 2 months time from 1991 to 2003.} \label{plot:average_degree_per_month}
\end{figure}
A downwards trend in the end for in-degree and upwards trend in the beginning are noticeable. Explanation to this phenomena is rather simple: older papers had more time to gain popularity and therefore have more incoming edges, whereas papers from 2003 do not have any papers in our dataset that cite them. The opposite happens with the out-degree. Earlier papers cited even earlier papers (from 1980s) that were not included in our graph and thus have little to no outgoing edges. It is also worth noticing that the growth of out-degree slows down closer to 2003. It means that almost enough papers are included in the graph at that time to publish a paper sourcing only these modern papers in our dataset.

\subsection{Clusterization}
Due to the duality of our problem, clusterization can be done either using node features (mainly papers' abstracts) or considering the graph structure (which is usually called community detection). As a measure of quality of the partition $P$ of graph $G$ we have chosen the metric modularity, which for directed graphs is defined as
$$\textbf{Modularity}(G, P) = \frac{1}{|E|} \sum_{u, v \in V} \left[\left(A_{u, v} - \frac{d_{\text{out}}(u)\cdot d_{\text{in}}(v)}{|E|}\right)\cdot \delta_P(u, v)\right],$$
where $A$  is the adjacency matrix and $\delta_P(u, v)$ is defined as
$$\delta_P(u, v) =
\begin{cases}
1, & \text{if nodes } u \text{ and } v \text{ lie in the same community in pratition } P \\
0, & \text{otherwise}
\end{cases}.
$$

First we did community detection on the graph. We compared Louvain and label propagation community detection algorithms since most other methods were extremely slow and thus not suitable for our graph. The best modularity was achieved by Louvain algorithm with partition into 31 classes with fixed seed compared to 969 classes using label propagation.

To do clusterixation we have divided the articles into the same number of clusters according to their abstracts using text embedding from `glove-wiki-gigaword-50` model (we averaged the embeddings of each word in the abstract) and KMeans algorithm. We also tried agglomerative clustering and Gaussian mixture. Modularity for all of them was really poor, which should not come as a surprise. Visualizations of resulting clustering can be found on Figures \ref{plot:clustering:agglomerative} and \ref{plot:clustering:gaussian_mixture}. The comparison of clustering algorithms can be seen in Table \ref{table:clusterization}.

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{Aglomerative.png}
\caption{Visualization of clustering using Agglomerative Clustering algorithm.}
\label{plot:clustering:agglomerative}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{gm.png}
\caption{Visualization of clustering using Gaussian Mixture model.}
\label{plot:clustering:gaussian_mixture}
\end{figure}

\begin{table}[h]
\centering
\begin{tabular}{c c c}
\hline\hline
Algorithm & Number of clusters & Modularity \\
\hline
Louvain & 31 & $0.655$ \\
Label Propagation & 969 & $0.541$ \\
Agglomerative Clustering & 31 & $-0.0002$ \\
Gaussian Mixture Model & 31 & $0.0008$ \\
\hline
\end{tabular}
\caption{Comparison of different clusterization algorithms.}
\label{table:clusterization}
\end{table}
